A demonstration of the \LaTeXe\ class file for \itshape{SAGE Publications}

### 2015-12-14 body/equilibrium.tex
Andrea's first complete review

\section{Heuristics for relevant contact selection}

\subsection{A heuristic for robust static equilibrium}
The planner is designed so that any contact configuration it outputs is in static equilibrium.
Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
that can result from perception and control uncertainties.

We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
From this value we can in turn either define a heuristic to choose the most robust candidate, or set a treshold on the required robustness
for any configuration. While the LP we propose is original, it is based on an analysis of the problem we proposed in \cite{Prete2016}, where the interested reader
can find more details.

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be express as a convex combination of its four generating rays.
Therefore we can express the non slipping constraint on $\mathbf{f}_i$ as:

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} , \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
\end{equation*}

we can stack the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.

From the Newton-Euler equations, and assuming a null acceleration, we also constrain the sum of contact forces to compensate the gravity at the center of mass:

@@

%~ Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
The right side of \ref{eq:new_eul} is independent from $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as:
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

If there exists a $\bm{\beta}^*$ satisfying equations \ref{eq:gen} and \ref{eq:new_eul2d}, it means that the configuration is statically balanced.
The problem can then be formulated as an LP:

@@

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
If $b_0$ is negative, the configuration is not in static equilibrium, and the value $b_0$ indicates "how far" from equilibrium
the configuration is. $b_0$ can thus be seen as a measure of robustness.

In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
after a trivial elimination of the Lagrange multipliers associated to the inequality constraints, the dual problem has a lower dimensionality:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

Indeed, from Slater's condition \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.

In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.
Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.

>>>>

% !TEX root =  ../main.tex
\section{Heuristics for relevant contact selection}

\subsection{A heuristic for robust static equilibrium}
The planner is designed so that any generated contact configuration is in static equilibrium.
Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real-world application, regardless of perception and control uncertainties.

We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
In turn, from this value we can either define a heuristic to choose the most robust candidate, or set a threshold on the required robustness\deleted{for any configuration}. While the presented LP is original, it is based on an analysis of the problem that we proposed in \cite{Prete2016}, where the interested reader can find more details.

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

According to Coulomb's law, \replaced{slipping occurs only if a contact force is on the boundaries of the friction cone.}{ the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.}
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be expressed as a convex combination of its four generating rays.
\deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} : \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i,
\end{equation*}
\added{where $\bm{\beta}_i$ contains the coefficient of the cone generators.}
We can then stack all the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta},
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = (\mathbf{f}_0,...,\mathbf{f}_e)$.

From the Newton-Euler equations, and assuming a null acceleration of the COM, we want the contact forces to compensate the gravity at the center of mass:

@@

%~ Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
The right hand side of \eqref{eq:new_eul} is independent of $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as: \adnote{It is not really necessary to show that equilibrium is independent of the com altitude here, anyway the com is given, not a variable, so it does not change the size of the problem}
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

If there exists a $\bm{\beta}^*$ satisfying \eqref{eq:gen} and \eqref{eq:new_eul2d}, it means that the configuration is statically balanced.
The problem can then be formulated as an LP:

@@

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \bm{1}\\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive then \eqref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
If $b_0$ is negative, the configuration is not in static equilibrium, and $b_0$ indicates ``how far'' from equilibrium the configuration is. We thus use $b_0$ as a measure of robustness.

In our implementation, rather than solving directly \eqref{eq:lin_prog_rob}, we solve an equivalent problem of smaller dimension that we get by taking the dual of \eqref{eq:lin_prog_rob} and eliminating the Lagrange multipliers associated to the inequality constraints:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

Indeed, from Slater's conditions \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal. Thus the optimal value of the LP~\eqref{eq:dual} is indeed the optimal $b_0$.

In conclusion, the LP~\eqref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.
Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.

### 2015-12-14 body/intro.tex
Andrea's first complete review

\subsection{Paper contribution and organization}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the \adnote{generation of the} guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties.}. We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
\item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact change, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
\end{itemize}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
For each problem, we introduce an original formulation of the related theoretical issues.
Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenario, or even better, for N robots and M scenarios}. The complete framework is implemented within
the open source platform HPP.

@@

To address \Pa we rigorously formulate a geometric condition for the true feasibility of a path, the reachability condition. We claim that this 
condition can be verified in a low-dimensional space, that we call $C_{reach}$. This formulation can make the planning of a guide trajectory more efficient computationally, while providing equivalent guarantees to planning directly in the configuration space.
In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained. This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot within a few milliseconds to a few seconds depending on the scenario.

To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain extremely rapidly \adnote{again, extremely rapidly} a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics.
In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
the Extended FORce Transmission ratio (EFORT).

>>>>

\subsection{Paper contribution and organization}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the \adnote{generation of the} guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must satisfy a property guaranteeing \adnote{replace 'satisfy a property guaranteeing' with 'guarantee'} the existence of a contact sequence to actuate it\adnote{Are you sure that you wanna use an 'endnote' here rather than a 'footnote'?}\endnote{This property is related to the controllability of the root actuated by the contact forces, but for discrete bounded properties. \adnote{not clear what 'discrete bounded properties' means}}. We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contacts\adnote{ forces?}, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
\item There is an infinite combination \adnote{There are infinite combinations} of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious contact sequences \adnote{repetition of 'contact sequence'}).
\end{itemize}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
For each problem, we introduce an original formulation of the related theoretical issues.
Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
The relaxation is justified empirically, with statistics presented for each robot and scenario presented\adnote{several robots and scenarios, or even better, for N robots and M scenarios}. The complete framework is implemented within
the open-source platform HPP.

@@

To address \Pa we rigorously formulate a geometric condition \replaced{that approximates}{for} the true feasibility of a path, the reachability condition. We claim that this 
condition can be verified in a low-dimensional space, which we call $C_{reach}$. This formulation can make the planning of a guide path more efficient computationally, while providing \replaced[remark={planning in $C_{reach}$ you don't account for equilibrium}]{simular}{equivalent} guarantees to planning directly in the configuration space.
In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained. This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot \replaced{in a timespan ranging from}{within} a few milliseconds to a few seconds, depending on the scenario.

To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theoretical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain extremely rapidly \adnote{again, extremely rapidly} a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, and ranked according to user-defined heuristics. \adnote{Not clear, first you say the candidates are sorted based on static equilibrium, then you say they are ranked based on user-defined heuristics. What's the difference between sorting and ranking?}
In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
the Extended FORce Transmission ratio (EFORT).

### 2015-12-14 body/overview.tex
Andrea's first complete review

%
\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {1} 
		\put (22,1) {2} 
		\put (42,1) {3} 
		\put (62,1) {4} 
		\put (83,1) {5} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. 1: Selection of reachable obstacles. 2: Entries of the limb samples database (with $N = 4$). 3: With a proximity query on the octree database, configurations too far from obstacles are eliminated. 4: The best candidate according to a user-defined heuristic $h$ is chosen. 5: The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}
\subsection{Computation of a guide path --- \Pa}
We first consider the problem of planning a relevant guide path. The objective is to compute a path of root placements which will allow contact creation. A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e. must lead to a valid sequence of contacts.

 An intuitive description of such placements is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1). We define formally $C_{reach}$, the set of interesting \adnote{'interesting' is vague, consider replacing with 'the set of root placements that allow the robot not to be in collision and be in contact with the environment'; alternatively you can even be more precise I write this in math, but maybe that's not the right place to be formal}  root placements, in which we compute a guide trajectory with a sampling based planner,  the Reachability-Based RRT --RB-RRT-- (Figure~\ref{fig:framework}--\Pa). Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
%
Details are presented in Section~\ref{rbprm}.

@@

The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb). The true feasibility of the input path is guaranteed by the first step.
This provides in turn a reasonable \adnote{FYI: you always wrote 'reasonable' as 'reasonnable'} guarantee that a contact configuration can be obtained for any root placement along the path. With a fixed root position, the dimensionality of the 
contact generation problem is reduced to the number of degrees of freedom of the considered limb. In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).

From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse-kinematics solver 
is used to maintain the contacts that were previously existing (TODO Figure). Often, these contacts cannot be maintained, because of joint limits or collisions (TODO Figure).
In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
The algorithm is designed so that only one contact can be created or broken between two successive configurations. This can be seen as a heuristic \adnote{why heuristic?} to ensure
that the interpolation between two configurations is achievable.
Details are presented in Section~\ref{sec:efort}.

To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, that have been
used in our experiments.

>>>>

%
\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {1} 
		\put (22,1) {2} 
		\put (42,1) {3} 
		\put (62,1) {4} 
		\put (83,1) {5} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. 1: Selection of reachable obstacles. 2: Entries of the limb samples database (with $N = 4$). 3: With a proximity query on the octree database, configurations too far from obstacles are eliminated. 4: The best candidate according to a user-defined heuristic $h$ is chosen. 5: The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}
\subsection{Computation of a guide path --- \Pa}
We first consider the problem of planning a relevant guide path. The objective is to compute a path of root placements that will allow contact creation. A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e. must lead to a valid sequence of contacts.

\added
{
More precisely, a root placement is truly feasible if and only if there exist at least one joint configuration such that the robot satisfies three conditions: }
\begin{description}
\item[C1] is not in collision with the environment, 
\item[C2] has at least one end-effector in contact with the environment, 
\item[C3] is in static equilibrium. 
\end{description}
\added{
The conditions C1 and C2 are purely geometrical and can be efficiently approximated even without computing the contact points. Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\cite{DelPrete2016}. Moreover our intuition suggests us that, for cluttered environment, C3 is not critical, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3. For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.
}

An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Fig.~\ref{fig:contact_gen}--1); not too close, because the robot must avoid collision (which is represented by the hull including the torso in Fig.~\ref{fig:contact_gen}--1). We define formally $C_{reach}$, the set of \replaced{root placements satisfying C1 and C2}{interesting root placements}, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Fig.~\ref{fig:framework}---\Pa). Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
%
Details are presented in Section~\ref{rbprm}.

@@

The second stage extends the guide path into a sequence of contact configurations (Fig.~\ref{fig:framework}--\Pb). \deleted{The true feasibility of the input path is guaranteed by the first step. This provides in turn a reasonable guarantee that a contact configuration can be obtained for any root placement along the path. }
With a fixed root position, the dimensionality of the 
contact generation problem is reduced to the number of degrees of freedom of the considered limb. In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).

From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root placement, an inverse-kinematics solver 
is used to maintain the contacts that were previously existing (TODO Figure). Often, these contacts cannot be maintained, because of joint limits or collisions (TODO Figure).
In such a case the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot (TODO FIGURE).
The algorithm is designed so that only one contact can be created or broken between two successive configurations. This \replaced{ensures}{can be seen as a heuristic to ensure} that the interpolation between two configurations is achievable.
Details are presented in Section~\ref{sec:efort}.

To select the ``most relevant'' contact configurations, user-defined heuristics can be chosen. In Section~\ref{heuristics} we present extensively two such heuristics, which have been used in our experiments.
